Rates of true polar wander in convecting planets

### 2016-01-21 tpw_rate.tex
Fix typo
Enormous derivation of magnitude of polar wander rate

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right)
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right]
\label{eq:liouville_ode}
\end{equation}
Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
If we enter the coordinate system of the convective moment of inertia $\mathbf{E}$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
\ifdetail

And here is the tedious algebra.  Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.  Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.  Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:
\begin{equation}
\begin{aligned}
 & \cos{\theta}\cos{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \dot{\phi} = \\
  &A \sin{\theta}\cos{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
 &\cos{\theta}\sin{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \dot{\phi} = \\
  &A \sin{\theta}\sin{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
 - &\sin{\theta} \dot{\theta} = \\
  &A \cos{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
\end{aligned}
\end{equation}
Now multiply the top equation by $\cos{\phi}$ and the second equation by $\sin{\phi}$ and add them together to get
\begin{equation}
\begin{aligned}
 & \cos{\theta}\cos^2{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \cos{\phi} \dot{\phi} + \cos{\theta}\sin^2{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \sin{\phi} \dot{\phi} = \\
  &A \sin{\theta}\cos^2{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] + \\
  &A \sin{\theta}\sin^2{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
 & \cos{\theta}\dot{\theta} =  
   A \sin{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
  &A \sin{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
\end{aligned}
\end{equation}
Now multiply the above by $\cos{\theta}$ and the third equation by $-\sin{\theta}$ and add them to get
\begin{equation}
\begin{aligned}
 & \cos^2{\theta}\dot{\theta} + \sin^2{\theta} \dot{\theta} =  
   A \sin{\theta} \cos{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
  &A \sin{\theta} \cos{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] - \\
  &A \cos{\theta} \sin{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
\end{aligned}
\end{equation}
After a lot of cancellation, we find
\begin{equation}
\begin{aligned}
 \dot{\theta} &= A \sin{\theta} \cos{\theta} \left[ - \lambda_3 + \lambda_1 \cos^2{\phi} + \lambda_2 \sin^2{\phi} \right] \\
              &= - A \sin{\theta} \cos{\theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
              &= -\frac{A}{2} \sin{2 \theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
\end{aligned}
\end{equation}
Likewise we may multiply the first equation by $-\sin{\phi}$ and the second equation by $\cos{\phi}$ and add them to get
\begin{equation}
\begin{aligned}
 & \sin{\theta} \dot{\phi} = A \sin{\theta}\cos{\phi}\sin{\phi} \left( \lambda_2 - \lambda_1 \right ) \\
& \dot{\phi} = \frac{A}{2} \sin{2 \phi} (\lambda_2-\lambda_1)
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
\label{eq:hydrostatic}
\end{aligned}
\end{equation}
Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2 \mathbf{v} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
\end{aligned}
\end{equation}

>>>>

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right)
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right]
\label{eq:liouville_ode}
\end{equation}
Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
If we enter the coordinate system of the convective moment of inertia $\mathbf{E}$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
\ifdetail

And here is the tedious algebra.  Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$.
We are interested in the rate of true polar wander, which is the angular change of the pole
per unit time. We can transform equation~\ref{eq:liouville_ode} into a scalar equation by dotting it
with $\dot{\mitbf{\omega}}$, since we are interested in the magnitude of the rate, rather than
the direction at the moment.

\begin{equation}
\begin{aligned}
\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} &= 
A^2 \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right]
\cdot \left[ \mathbf{E}\cdot \mitbf{\omega} - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)\mitbf{\omega}\right] \\
\dot{\mitbf{\omega}}^2  &= 
A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)\cdot \left(\mathbf{E}\cdot \mitbf{\omega}\right)
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right) \left( \mitbf{\omega}\cdot \mathbf{E} \cdot \mitbf{\omega} \right) 
 + \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 \mitbf{\omega}\cdot \mitbf{\omega} 
\right]\\
\dot{\mitbf{\omega}}^2  &= 
A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 
\right]\\
\end{aligned}
\end{equation}
Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
It is most convenient to choose the principal axes of $\mathbf{E}$, where
\begin{equation}
\mathbf{E} = 
\begin{bmatrix}
\lambda_1 & 0 & 0 \\
0 & \lambda_2 & 0 \\
0 & 0 & \lambda_3 \\
\end{bmatrix}
\end{equation}
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta} \cos{\phi} \\
\sin{\theta} \sin{\phi} \\
\cos{\theta} \\
\end{bmatrix}
\end{equation}

As a warmup, consider the case where $\phi = 0$, then  
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta}  \\
0 \\
\cos{\theta} \\
\end{bmatrix}
\end{equation}
We can evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \\
0 \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

\begin{equation}
\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
\end{equation}

\begin{equation}
\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
\lambda_1 \sin^2{\theta} + \lambda_3 \cos^2{\theta}
\end{equation}

\begin{equation}
\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
\end{equation}
Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
\begin{equation}
\begin{aligned}
\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
 \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
 &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
 - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
 &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
\end{aligned}
\end{equation}

Now consider the more general case whhere $\phi \ne 0$.
Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \cos{\phi}\\
\lambda_2 \sin{\theta} \sin{\phi} \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

\begin{equation}
\left(\mathbf{E}\cdot \mitbf{\omega}\right)^2 = \lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta}
\end{equation}

\begin{equation}
\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} = 
\lambda_1 \sin^2{\theta} \cos^2{\phi} + \lambda_2 \sin^2{\theta} \sin^2{\phi} + \lambda_3 \cos^2{\theta}
\end{equation}

\begin{equation}
\begin{aligned}
\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
&\lambda_1^2 \sin^4{\theta} \cos^4{\phi} + 
\lambda_2^2 \sin^4{\theta} \sin^4{\phi} +
\lambda_3^2 \cos^4{\theta} \\
&+ 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
&+ 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
&+ 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
\end{aligned}
\end{equation}
Again, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
\begin{equation}
\begin{aligned}
\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
 \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
&=
\lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta} \\
&-\lambda_1^2 \sin^4{\theta} \cos^4{\phi} - 
\lambda_2^2 \sin^4{\theta} \sin^4{\phi} -
\lambda_3^2 \cos^4{\theta} \\
&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
&=
\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
&- \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}(1-\sin^2{\phi}) + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}(1-\cos^2{\phi}) \right] \\
&+ \lambda_3^2 \cos^2{\theta}(1-\cos^2{\theta}) \\
&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
&=
\sin^2{\theta} \left[\lambda_1^2 \cos^2{\phi} (1 - \sin^2{\theta}) + \lambda_2^2 \sin^2{\phi} (1-\sin^2{\theta}) \right] \\
&+ \sin^2{\theta} \left[ \lambda_1^2 \sin^2{\theta} \cos^2{\phi}\sin^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi}\cos^2{\phi} \right] \\
&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
&=
\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} \right] \\
&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left[ \lambda_1^2 + \lambda_2^2 -2 \lambda_1\lambda_2 \right] \\
&+ \lambda_3^2 \cos^2{\theta}\sin^2{\theta} \\
&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
&=
\sin^2{\theta}\cos^2{\theta} \left[\lambda_1^2 \cos^2{\phi} + \lambda_2^2 \sin^2{\phi} + \lambda_3^2 - 2 \lambda_1 \lambda_3 \cos^2{\phi} - 2 \lambda_2 \lambda_3 \sin^2{\phi} \right] \\
&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
&=
\sin^2{\theta}\cos^2{\theta} \left[( \lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \cos^2{\phi} + (\lambda_2^2 + \lambda_3^2 - 2 \lambda_2 \lambda_3 ) \sin^2{\phi} \right] \\
&+ \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
&=
\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] \\
&+ \sin^2{\theta} \cos^2{\theta} \tan^2{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \\
&=
\sin^2{\theta}\cos^2{\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi}
+ \tan^2{\theta} \cos^2{\phi}\sin^2{\phi} \left( \lambda_2 - \lambda_1 \right)^2 \right] \\
\end{aligned}
\end{equation}

At long last, we get
\begin{equation}
\begin{aligned}
\frac{\dot{\mitbf{\omega}}^2}{A^2}  &= 
\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi}
+ \frac{1}{4}\tan^2{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \right] \\
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
\label{eq:hydrostatic}
\end{aligned}
\end{equation}
Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
\end{aligned}
\end{equation}
